The Courtroom Challenge—A Conjecture on the Unavoidable Cost of LLM Verification

### 2025-06-23 paper/main.tex
Add files via upload

\subsection{Information–theoretic lower bound}
\label{sec:info-bound}

\begin{lemma}[Interactive-Fano corollary]\label{lem:ErrLower}
  Fix a validator $V$ run for $T$ rounds under margin parameter $\delta>0$.
  Let
  \[
    \kappabar
      :=\mathbb{E}_{\omega}\bigl[\kappa(\omega)\bigr],
    \qquad
    \Err_{\mathrm{bar}}
      :=\mathbb{E}_{\omega}\bigl[\Err(\omega)\bigr].
  \]
  Suppose the transcript $X$ satisfies
  $I(\theta;X)\le I$
  under a uniform prior on an $M$-point label set ($M\ge 2$).
  Then
  \begin{equation}\label{eq:ErrLower}   % <-- new label
    \boxed{%
      \Err_{\mathrm{bar}}
      \;\;\ge\;\;
      \frac{\delta\,\kappabar\,T}{4}\;
      \Bigl(
        1-\dfrac{I+\log 2}{\log M}
      \Bigr)
    }.
  \end{equation}
\end{lemma}

\begin{proof}[Sketch]

\textbf{Classical identification bound.}
Choosing the 0–1 loss $L(M,X)=\mathbf 1_{\{\hat M(X)\neq M\}}$
and threshold $\Delta=\tfrac12$
in Proposition 3 of \citet{chen2024interfano} yields
\[
  P_{\mathrm{id}}
  \;\ge\;
  \tfrac12\Bigl(1-\tfrac{I+\log 2}{\log M}\Bigr).
\]

\medskip
\noindent
To obtain a bound on label error we apply the same proposition with 
threshold $\Delta=\tfrac12$ to the indicator
\[
  L^\star \;=\; \mathbf 1_{\{\Err>\Delta_{\mathrm{Err}}\}},
  \qquad
  \Delta_{\mathrm{Err}} := \tfrac12\,\delta\,\kappabar\,T.
\]

\item
  Because $\log \!\bigl(1/\rho_{1/2}\bigr)\ge\log M$ for the uniform
  prior, the proposition yields the lower bound
  \[
    \Prb\!\bigl[\Err>\Delta_{\mathrm{Err}}\bigr]
      \;\ge\;
      \frac12\Bigl(
        1-\frac{I+\log 2}{\log M}
      \Bigr).
  \]

\item
  Multiply this probability by
  $\Delta_{\mathrm{Err}}=\tfrac12\,\delta\,\kappabar\,T$
  to bound the expectation:
  \[
    \E[\Err]
      \;\ge\;
      \Delta_{\mathrm{Err}}\;
      \Prb[\Err>\Delta_{\mathrm{Err}}]
      \;=\;
      \frac{\delta\,\kappabar\,T}{4}\;
      \Bigl(
        1-\frac{I+\log 2}{\log M}
      \Bigr).
      \qedhere
  \]

@@

\paragraph{Discussion.}
Inequality~\eqref{eq:ErrLower} separates the lower bound into two
factors:

\begin{enumerate}[label=(\roman*)]
\item the \emph{size factor} $\displaystyle\frac{\delta\,\kappabar\,T}{4}$,
      which grows linearly with the horizon~$T$ and is large whenever
      the validator answers many queries ($\kappabar$ moderate) under a
      tight margin~$\delta$;
\item the \emph{information gap}
      $\displaystyle\Bigl(1-\tfrac{I+\log 2}{\log M}\Bigr)$,
      which is positive as soon as the transcript carries fewer bits
      than the label space can encode.
\end{enumerate}
Together they formalise our narrative: \emph{a low-cost, low-friction
validator cannot keep cumulative label error small indefinitely as the
stream length~$T$ grows}, unless it either (a) escalates a non-negligible
fraction of rounds (making $\kappabar$ small and thus incurring cost) or
(b) relaxes the margin~$\delta$.  In particular, when $M$ is
exponentially large and $I$ is bounded by design, $\Err_{\mathrm{bar}}$
grows linearly with~$T$.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Setup}\label{sec:setup}

\paragraph{Per-round notation.}
Fix a margin parameter $\delta\in(0,1]$.
For every run $\omega$ and round $t=1,\dots,T$ set
\begin{align*}
  y_t      &:=y_t(\theta)          &&\text{(ground-truth label)},\\
  h_t      &:=h_t(\omega)          &&\text{(validator’s automatic prediction)},\\
  \mathrm{Esc}_t&:=\mathbf 1\{\text{round $t$ is escalated}\}.
\end{align*}
Define the \emph{leverage rounds}
\begin{equation}\label{eq:kappaDef}
  \mathcal L(\omega):=\Bigl\{\,t:\,
     \mathrm{Esc}_t=0\;\text{and}\;|y_t-h_t|\ge\delta\Bigr\},
  \qquad
  \kappa(\omega):=\frac{|\mathcal L(\omega)|}{T}.
\end{equation}
The cumulative label error on those rounds is
\begin{equation}\label{eq:ErrDef}
  \Err(\omega):=\sum_{t\in\mathcal L(\omega)}
                \mathbf 1\!\bigl[h_t\neq y_t\bigr].
\end{equation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Information-theoretic lower bound}\label{sec:info-bound}

\begin{lemma}[Interactive-Fano corollary]\label{lem:ErrLower}
Let $V$ be any validator run for $T$ rounds.
With $\kappa(\omega)$ and $\Err(\omega)$ defined in
\eqref{eq:kappaDef}–\eqref{eq:ErrDef}, write
\[
  \bar\kappa:=\E[\kappa],
  \qquad
  \Err_{\mathrm{bar}}:=\E[\Err].
\]
Assume the transcript $X$ obeys the information budget
\(I(\theta;X)\le I\) under the \emph{uniform} prior on an $M$-point
label set ($M\ge2$).  Then
\begin{equation}\label{eq:ErrLower-new}
  \boxed{\;
    \Err_{\mathrm{bar}}
      \;\ge\;
      \frac{\delta\,T}{2}\;
      \Bigl(\bar\kappa+\gamma-1\Bigr),
      \quad
      \gamma:=\frac12\Bigl(1-\tfrac{I+\log 2}{\log M}\Bigr)
  }.
\end{equation}
In particular, the bound is positive—and linear in $T$—whenever the
validator escalates fewer than a $(1-\gamma)$-fraction of rounds
\((\bar\kappa>1-\gamma)\).
\end{lemma}

\begin{proof}[Sketch]
Step 1 (separation).
For any $m\neq m'$ and transcript $x$ at most one of
$y_t(m),y_t(m')$ can equal $h_t$ on a leverage round, hence
\[
  \Err_m(x)+\Err_{m'}(x)\;\ge\;\delta\,\kappa(\omega)\,T.
  \tag{$\ast$}
\]

\noindent
Step 2 (interactive Fano). 
Set the path-dependent loss
\(L^\star(m,x):=\mathbf 1\!\bigl[\Err_m(x)>
   \tfrac12\delta\kappa(\omega)T\bigr]\)
and $\Delta=\tfrac12$ in Proposition 3 of
\citet{chen2024interfano}.
Inequality ($\ast$) fulfils the separation requirement
\citep[Eq.\,(5)]{chen2024interfano}; thus
\[
  \Pr\!\bigl[\Err(\theta,X)>
     \tfrac12\delta\kappa(\omega)T\bigr]
  \;\ge\;\gamma.
  \tag{$\dagger$}
\]

\medskip\noindent
Step 3 (tail $\Rightarrow$ expectation, expanded).
With
\[
  \mathcal E:=\Bigl\{\Err>\tfrac12\delta\,\kappa T\Bigr\},
  \qquad
  \kappa_{\!*}:=\E[\kappa\mid\mathcal E],
\]
apply the basic fact $\E[Z]\ge a\,\Pr(Z>a)$ to
\(Z=\Err\) and \(a=\tfrac12\delta\,\kappa T\):
\[
\begin{aligned}
  \Err_{\mathrm{bar}}
    &=\E[\Err] \\[2pt]
    &\ge
      \E\!\Bigl[
        \tfrac12\delta\,\kappa T\;
        \mathbf 1_{\{\Err>\tfrac12\delta\,\kappa T\}}
      \Bigr] \\[6pt]
    &=\frac{\delta T}{2}\,
      \E\!\bigl[\kappa\,\mathbf 1_{\mathcal E}\bigr] \\[6pt]
    &=\frac{\delta T}{2}\,
      \kappa_{\!*}\;
      \Pr(\mathcal E).
\end{aligned}
\]

@@

\paragraph{Discussion.}
Write the bound as  
\(\Err_{\mathrm{bar}}\!\ge(\delta T/2)\,\max\{0,\bar\kappa+\gamma-1\}\).
Because \(\gamma>0\) whenever the transcript leaks
fewer than \(\log M-\log2\) bits, any validator that
handles a \emph{non-trivial} share of rounds automatically
\((\bar\kappa>1-\gamma)\)
faces a \emph{linear-in-$T$} unavoidable error floor.
Escalating a larger fraction of rounds (making $\bar\kappa$ smaller) or
relaxing the margin $\delta$ are the only ways to suppress this growth,
precisely matching the “friction buys accuracy’’ narrative.
